From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-23 appendixB/appendixB.Rmd
WIP
van Kampen

which for historical reasons is known as a master equation. From this we can directly calculate the mean and variance for this process by multiplying by $n$ or $n^2$ and summing over all $n$. These calculations are greatly simplified by observing the following property of the step operator: for any pair of test functions $f_n$, $g_n$, 
\begin{align}
\sum_{n=0}^{N-1}  g_n \mathbb E f_n = \sum_{n=1}^N f_n \mathbb E^{-1} g_n \nonumber
\end{align}

@@

For small $\epsilon$ we can ignore the terms of $\mathcal O(\epsilon^2)$, and we have recovered the diffusion approximation.  Unfortunately, we have not been very precise about what we mean by $\epsilon$ -- nature gives us no such obvious small parameter that we can use to "scale down" step sizes and "scale up" rates. The van Kampen expansion makes this notion more precise by introducing the concept of system size.  Instead of counts we consider a change of units into counts per system size, e.g. population density.  We can imagine drawing ever larger rings in our landscape of individuals, defining an ever bigger and bigger "system."

@@

>>>>

which for historical reasons [see @vanKampen2007] is known as a master equation. From this we can directly calculate the mean and variance for this process by multiplying by $n$ or $n^2$ and summing over all $n$. These calculations are greatly simplified by observing the following property of the step operator: for any pair of test functions $f_n$, $g_n$, 
\begin{align}
\sum_{n=0}^{N-1}  g_n \mathbb E f_n = \sum_{n=1}^N f_n \mathbb E^{-1} g_n \nonumber
\end{align}

@@

For small $\epsilon$ we can ignore the terms of $\mathcal O(\epsilon^2)$, and we have recovered the diffusion approximation.  As noted above, we can write this PDE in the more compact notation of an SDE, substituting in our original terms for birth and death:

\begin{equation}
\ud X_t = \left[ b(x) - d(x) \right] \ud t + \sqrt{b(X_t) + d(X_t)} \ud W_t
\end{equation}

Unfortunately, we have not been very precise about what we mean by $\epsilon$ -- nature gives us no such obvious small parameter that we can use to "scale down" step sizes and "scale up" rates. The van Kampen expansion makes this notion more precise by introducing the concept of system size.  Instead of counts we consider a change of units into counts per system size, e.g. population density.  We can imagine drawing ever larger rings in our landscape of individuals, defining an ever bigger and bigger "system."

@@

The van Kampen system size expansion, shown for a general birth-death
process in Appendix B, derives an *ordinary* differential equation (ODE)
for the macroscopic process, coupled to a *linear* SDE which
governs the *deviations* ($\xi$)  from the macroscopic equation:

\begin{align}
\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) + \mathcal{O}(N^{-1}) \\
\mathrm{d} \xi &=  \partial_{x}  \left( b(x)  - d(x) \right)  \xi \mathrm{d}t + \sqrt{b(x) + d(x)} \mathrm{d} B_t   + \mathcal{O}(N^{-\tfrac{1}{2}}
\end{align}

### 2018-02-23 paper/paper.Rmd
WIP
van Kampen

@@

$$ \textrm{d} N_t = \underbrace{f(N_t) \textrm{d} t}_{\textrm{det skeleton}} + \underbrace{\sigma_d \sqrt{N_t} \phantom{\cdot}\textrm{d}B_t^{(d)} }_{\textrm{demographic noise}} + \underbrace{\sigma_e N_t \phantom{\cdot}\textrm{d}B_t^{(e)}}_{\textrm{environmental noise}}$$

@@

Before we follow that path, I will highlight

I will review how a shift towards more mechanistic descriptions of the origins of noise provide a more nuianced view of when the canonical model does an does not hold in the context of both demographic and environmental stochasticity.  My approach will highlight two fundamental tools: an ability for efficient, exact simulation using @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen1976].

which is rooted in the mathematical theorems of Thomas Kurtz [@Kurtz1970 & @Kurtz1971, @Kurtz1972 @Kurtz1978], in place of the more popular diffusion approximation derived by Taylor series [e.g. @Nisbet1982, @Lande2003; rooted in the earlier work of @Kramers1940 and @Moyal1949, see comparisons in @vanKampen2007 and @Gardiner2009].

@@

Demographic stochasticity refers to fluctuations in population sizes
or densities that arise from the fundamentally discrete nature of
individual birth and death events.  Demographic stochasticity is a
particularly instructive case for illustrating a mechanism for how
noise arises as an aggregate description from a lower-level mechanistic
process.  We summarize the myriad lower-level processes that
mechanistically lead to the event of a 'birth' in the population
as a probability: In a population of $N$ identical individuals at
time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate
that can depend on the population size, $N$), which increases the
population size to $N+1$. Similarly, death events occur with probability
$d_t(N_t)$, decreasing the population size by one individual, to $N-1$.  Assuming
each of these events are independent, this is a state-dependent
Poisson process. The change in the probability of being in state
$N$ is given by the sum over the ways to enter the state, minus the
ways to leave the state: a simple expression of probability balance
known as the "master equation" [@vanKampen2007].

The @Gillespie1977 provides a simple and exact algorithm
for simulating demographic stochasticity at an individual level.  
Free from both the approximations and mathematical complexity,
the Gillespie algorithm offers a simple and powerful approach to simulating
stochastic phenomena that remains underutilized.

We can illustrate both the Gillespie algorithm and the van Kampen
system size expansion using @Levins1969 classic patch model,

@@

where $n$ individuals compete for a
finite number of suitable habitats $N$.  Individuals die a constant
rate $e$, and produce offspring at a constant rate $c$ who then
have a probability of colonizing an open patch that is simply
proportional to the fraction of available patches, $1 - n/N$.  
We choose this model because it has precisely same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  As we shall see, this is not merely a notational
convience but leads to real differences in observed stochastic properties
of the model.  Though we focus on this simple model for clarity of 
presentation, both computational and mathematical framework presented here 
can easily be applied to alternative models to reach the same conclusions.

Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$.

@@

The van Kampen system size expansion, shown for a general birth-death
process in Appendix B, derives an *ordinary* differential equation (ODE)
for the macroscopic process, coupled to a *linear* SDE which
governs the *deviations* ($\xi$)  from the macroscopic equation:

\begin{align}
\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) + \mathcal{O}(N^{-1}) \\
\mathrm{d} \xi &=  \partial_{x}  \left( b(x)  - d(x) \right)  \xi \mathrm{d}t + \sqrt{b(x) + d(x)} \mathrm{d} B_t   + \mathcal{O}(N^{-\tfrac{1}{2}}
\end{align}

\begin{align}
\frac{\mathrm{d} x }{\mathrm{d} t}  &= b(x) - d(x) + \mathcal{O}(N^{-1})\\
\frac{\mathrm{d} \sigma^2 }{\mathrm{d} t}  &= 2 \left( \partial_x b(x)  - \partial_x d(x) \right) \sigma^2 + b(x) + d(x)  + \mathcal{O}(N^{-\tfrac{1}{2}}) \label{bd}
\end{align}

The dynamics for the average population
match the population-level equation.^[up to fluctuations of order
$\tfrac{1}{N}$. We can actually both observe and successfully
approximate this higher-order correction from additional terms in
the system size expansion to recover the phenomenon of stochastic
inflation [@Gonzalez2002], where the average population size can
be larger or smaller than predicted by the deterministic model --
see Appendix A for examples in simulations compared to predictions
of the system size expansion].

Note that the van Kampen expansion is not identical to the SDE 
postulated by the-so called canonical equation.  
As shown in Appendix B, the popular expansion

The change in variance can be seen
as a balance between the stochastic contribution, $b(x) + d(x)$
and the deterministic contribution, $\partial_x \left(b(x) - d(x)\right)$. 
The deterministic contribution is the slope of the macroscopic
equation, that is, the eigenvalue indicating the stability of the
equilibrium point.  For stable points, this value must be negative
and precisely balance the strictly positive stochastic contribution 
of $b(x) + d(x)$ (the greater the rate of stochastic events involved, 
the greater the fluctuations).  This balance at steady-state is known as the
fluctuation-dissipation theorem [@vanKampen2007], an will prove a useful
expression in understanding stochastic dynamics.  Because Eq \eqref{bd}
also describes the time-evolution of these fluctuations, we can similarly
solve for the correlation between two points separated by time $\tau$.

>>>>

<!-- FIXME look at Coulson, Bjornstead, Ovaskeinen and others and see how they define scope to focus on "population ecology" -->

@@

\begin{equation}
\textrm{d} N_t = \underbrace{f(N_t) \textrm{d} t}_{\textrm{det skeleton}} + \underbrace{\sigma_d \sqrt{N_t} \phantom{\cdot}\textrm{d}B_t^{(d)} }_{\textrm{demographic noise}} + \underbrace{\sigma_e N_t \phantom{\cdot}\textrm{d}B_t^{(e)}}_{\textrm{environmental noise}} 
\label{canonical}
\end{equation}

@@

Before we follow that path, we will examine how this canonical model can emerge from a lower-level description and how it can deviate from that description. This approach will also set the stage to see how our understanding of stochastic models can be extended to better reflect natural processes that shape not only for the so-called deterministic skeleton, but the noise as well.  In so doing, I will highlight two tools fundamental to this analysis, one computational and one analytical: the exact simulation method of @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen1976].

@@

Demographic stochasticity refers to fluctuations in population sizes
or densities that arise from the fundamentally discrete nature of
individual birth and death events.  Demographic stochasticity is a
particularly instructive case for illustrating a mechanism for how
noise arises as an aggregate description from a lower-level mechanistic
process.  We summarize the myriad lower-level processes that
mechanistically lead to the event of a 'birth' in the population
as a probability: In a population of $N$ identical individuals at
time $t$, a birth occurs with probability $b_t(N_t)$ (_i.e._ a rate
that can depend on the population size, $N$), which increases the
population size to $N+1$. Similarly, death events occur with probability
$d_t(N_t)$, decreasing the population size by one individual, to $N-1$.  Assuming
each of these events are independent, this is a state-dependent
Poisson process. The change in the probability of being in state
$N$ is given by the sum over the ways to enter the state, minus the
ways to leave the state: a simple expression of probability balance
known as the "master equation" [@vanKampen2007].  Note that in general
this approach is equally applicable to stochastic transitions of any sort,
not just step sizes of +/- 1 and not just birth and death events, but
can include transitions between stage classes or trait values, including
mutations to continuously-valued traits in evolutionary dynamics [e.g. @Boettiger2009].

The @Gillespie1977 provides an exact algorithm
for simulating demographic stochasticity at an individual level.  
The algorithm is merely a simple and direct implemtentation of the master
equation, progressing in random step sizes deterimed
by the waiting time until the next event.
Free from both the approximations and mathematical complexity,
the Gillespie algorithm is an interesting example of where we rely on a
numerical implementation to check the accuracy of an analytic approximation,
even in the case of simple models such as we will discuss.  Though the 
algorithm is often maligned as numerically demanding, it can be run much
more effectively even on large models on today's computers than when
it was first developed in the 70s, and remains an underutilized approach
for writing simple and approximation-free[^2] stochastic ecological models.

[^2]: that is, free from the approximation made by SDE models as we see in the
van Kampen example.  All models are, of course, only approixmations.

As our objective is to tie the origins of noise more closely to biological
processes, it will be helpful to make the notion of a master equation concrete
with a spefiic example.  We will focus on the classic case of @Levins1969 patch model,
to illustrate the Gillespie algorithm and the van Kampen system size expansion

@@

where $n$ individuals compete for a
finite number of suitable habitats $N$.  Individuals die a constant
rate $e$, and produce offspring at a constant rate $c$ who then
have a probability of colonizing an open patch that is simply
proportional to the fraction of available patches, $1 - n/N$.

We choose this model because it has precisely same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  As we shall see, this is not merely a notational
convience but leads to real differences in observed stochastic properties
of the model. Both the numerical approach (details in Appendix A) and the 
analytical approach (Appendix B) illustrate that it is straight-forward
to use this same approach in alternate formuluations.

Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$, illustrating the magnitude of these fluctuations indeed scales in proportion
to $\sqrt{N}$ as postulated by the canonical equation, Eq \eqref{canoical}.

@@

The van Kampen system size expansion provides a mechanism to approximate the individual
dynamics represented by the master equation (equivalently, by the Gillespie simulation)
with a population-level description of the process.  This is important not only because
it will illustrate both origins and approximations that lie behind the population level
models that predominate this literature, but will give our syntheses a chance to point
out some more subtle misconceptions that have also arisen. Details of this expansion
for a generic birth death process and comparison with the more familiar diffusion 
approximation are found in Appendix B.

van Kampen's approximation follows a central-limit theorem style argument established rigorously in
the theorems of Kurtz at about the same time [@Kurtz1970 & @Kurtz1971, @Kurtz1972, in particular, see Thereom 4.4. @Kurtz1978].
The heart of the approximation is to imagine a population on some large landscape of which
we sample only a fixed area to estimate the population density.  That area is our 
system size, which as it expands includes more and more total individuals and
experiences proportionally smaller fluctionations in population density.  This 
results in the conclusion that the population density obeys a simple Gaussian
distribution, whose mean size $x$ and variance  $\sigma^2$ are given by *ordinary* differential equations:

\begin{align}
\frac{\mathrm{d} x }{\mathrm{d} t}  &= b(x) - d(x) + \mathcal{O}(N^{-1})\\
\frac{\mathrm{d} \sigma^2 }{\mathrm{d} t}  &= 2 \partial_x  \left[ b(x)  - d(x) \right] \sigma^2 + b(x) + d(x)  + \mathcal{O}(N^{-\tfrac{1}{2}}) \label{vanKampen}
\end{align}

This result is striking for a variety of reasons.  First, we see the recovery of the 
"deterministic skeleton," in which we naively replace population density $x/N$ for
population count $n$ describes the *expected* density of the population
(i.e. averaged over an esemble of replicate realizations of the Gillespie simulations),
and with remarkable accuracy, up to order $1/N}$, i.e. a single individual.^[We can actually extend the van Kampen expansion to higher order and successfully approximate even this higher-order correction to the mean dynamics, recovering the phenomenon of stochastic inflation [@Gonzalez2002], where the average population size can be larger or smaller than predicted by the deterministic model depending on the curvature of the macroscopic equation].
The variance term is equally interesting.  First, we note that noise arises from
the interaction of individual birth and death events that defined our process,
illustrating the origins of this noise.  Because both individual events of births
and deaths add randomness, they both add to the growth of varation observed, i.e.
$b(x) + d(x)$.  Second, we see that dynamics for the variance $\sigma_^2$ are *non-linear* whenever
$b(x)$ or $d(x)$ are non-linear, so that different values of the population density
will experience different noise intensities! In fact, this variance can only
have an equilibrium size if the macroscopic equation has an equilibrium ($\partial_x  \left[ b(x)  - d(x) \right] = 0$)
Third, this expansion shows the variance
equation is also less precise, accurate only to order $1/\sqrt{N}$.

In the interests of a theoretical synthesis, I observe that this result
is *different* than the canonical equation, Eq \eqref{canonical}, or more generally
the diffusion approximation [or Kramer-Moyal expansion @Kramers1940; @Moyal1949; @Gardiner2009]
which is much more typical in the ecological literature [e.g. the books of @Nisbet1982 and @Lande2003
or the review in @Ovaskeinen2010]. @Black2012 gives an excellent introduction 
to the van Kampen expansion as a connection between individual-based and population-level
model descriptions and applications, but fails to point out these differences,
mistakenly characterizing the van Kampen expansion as synonomous with the diffusion approximation.
In brief, the diffusion approximation does not lead to a Gaussian distribution
(equivalently, an SDE or PDE with linear noise term, see Appendix B), but rather
a *non-linear* SDE of the form (e.g. @Ovaskeinen2010)

\begin{equation}
\ud X_t = \left[ b(x) - d(x) \right] \ud t + \sqrt{b(X_t) + d(X_t)} \ud W_t \label{diffusion}
\end{equation}

This result has several potential draw-backs.  First, as noted in @vanKampen2007 or
@Gardiner2009 (see Appendix B), the Kramer-Moyal expansion approach rests on rather
a more dubious double limit that often has no clear biological intepretation and
can restrict its validity (though I believe theorem 3.3 of @Kurtz1978 establishes
this approximation on firmer grounds).  Second, the non-linear SDE formulation can
prove more difficult to work with than the convenient ODE expressions from van Kampen,
Eq \eqref{\vanKampen} (which can be expressed as a coupled ODE for the mean and 
a linear SDE for deviations).  Third, while both approximations break down
in the case of large deviations [see @Ovaskeinen2010], the van Kampen expansion
shows that the macroscopic equation will continue to hold even when the 
variance does not; in contrast to the non-linear SDE from the diffusion approximation.

- Link back to Levins, show where same
- IMPORTANCE of fluctuation dissipation theorem to the tranisition probability discussion!!
- ENV noise and colored noise

The change in variance can be seen
as a balance between the stochastic contribution, $b(x) + d(x)$
and the deterministic contribution, $\partial_x \left(b(x) - d(x)\right)$. 
The deterministic contribution is the slope of the macroscopic
equation, that is, the eigenvalue indicating the stability of the
equilibrium point.  For stable points, this value must be negative
and precisely balance the strictly positive stochastic contribution 
of $b(x) + d(x)$ (the greater the rate of stochastic events involved, 
the greater the fluctuations).  This balance at steady-state is known as the
fluctuation-dissipation theorem [@vanKampen2007], an will prove a useful
expression in understanding stochastic dynamics.  Because Eq \eqref{vanKampen}
also describes the time-evolution of these fluctuations, we can similarly
solve for the correlation between two points separated by time $\tau$.
